---
output:
  pdf_document: default
  html_document: default
---

```{r}
library(dplyr)
library(tidyr)
library(brms)
library(rstan)
```

# Load data

```{r}
# 1. Load Data
# ------------------------------------
raw_data <- read.csv("../data/coded_output.csv", stringsAsFactors = FALSE)

# Create a lookup for a representative lemma (shortest string) per lemma_id
lemma_lookup <- raw_data %>%
  filter(!is.na(lemma), lemma != "") %>%
  group_by(lemma_id) %>%
  slice_min(nchar(lemma), n = 1, with_ties = FALSE) %>%
  select(lemma_id, lemma_rep = lemma) %>%
  ungroup()

raw_data <- raw_data %>%
  left_join(lemma_lookup, by = "lemma_id") %>%
  mutate(lemma = lemma_rep) %>%
  select(-lemma_rep)

model_data <- raw_data %>%
  # A. Filter: Keep only rows where is_bipartite is coded (not NA)
  filter(!is.na(is_bipartite)) %>%
  
  # B. Filter: Drop rows where BOTH leveling indicators are missing
  #    (Keep row if at least one of them has a value)
  filter(!is.na(is_leveled_vowel_pres) | !is.na(is_leveled_cons_pres))


# 2. Preprocessing Pipeline
# ------------------------------------
model_data <- raw_data %>%
  # A. Filter: Keep only rows where is_bipartite is coded (not NA)
  filter(!is.na(is_bipartite)) %>%
  
  # B. Filter: Drop rows where BOTH leveling indicators are missing
  #    (Keep row if at least one of them has a value)
  filter(!is.na(is_leveled_vowel_pres) | !is.na(is_leveled_cons_pres)) %>%
  
  # C. Transformation: Pivot to Long Format
  #    We want separate rows for Vowel leveling events and Consonant leveling events
  pivot_longer(
    cols = c(is_leveled_vowel_pres, is_leveled_cons_pres),
    names_to = "element_type_raw",
    values_to = "has_levelled"
  ) %>%
  
  # D. Post-Pivot Cleaning
  #    Remove rows created by the pivot that are NAs (e.g. if a verb only had vowel info)
  filter(!is.na(has_levelled)) %>%
  mutate(
    # Clean up element type names
    element_type = if_else(element_type_raw == "is_leveled_vowel_pres", "vowel", "consonant"),
    
    # Ensure factors
    element_type = as.factor(element_type),
    is_bipartite = as.factor(is_bipartite),
    variety = as.factor(variety),
    lemma_std = as.factor(lemma_id), # Assuming 'lemma' column is the standard ID

    id = as.factor(id),           # Document ID
    std_infl = as.factor(std_infl),
    
    # Log transform frequency (adding small constant to avoid log(0) if needed)
    log_freq = log(lemma_freq_per_1000 + 0.0001)
  ) %>%
  
  # Select final columns for cleanliness
  select(lemma, lemma_std, date, id, variety, std_infl, 
         log_freq, is_bipartite, element_type, has_levelled)

model_data <- unique(model_data)

head(model_data)

```

# Model comparison

```{r}
library(loo)

# 1. Load the models from the models folder
fit_no_slopes <- readRDS("models/comprehensive_brms_fit_no_slopes.rds")
fit_k3        <- readRDS("models/comprehensive_brms_fit_k3.rds")
fit_full      <- readRDS("models/comprehensive_brms_fit.rds")
linear_freq   <- readRDS("models/linear_freq_brms_fit_k3.rds")
refined_fit_k4 <- readRDS("models/refined_model_brms_fit_k4.rds")
linear_freq_refined_k4 <- readRDS("models/refined_model_brms_fit_k4_linear_freq.rds")


# 2. Add LOO criteria to each model
# We use moment_match = TRUE if there are high Pareto k values, 
# though this requires the original data and environment.


# 3. Compare the models
# The comparison will show the difference in Expected Log Predictive Density (elpd_diff).
# The model with elpd_diff = 0 is the best-performing model.
comparison <- loo_compare(fit_no_slopes, fit_k3, fit_full, linear_freq, refined_fit_k4, linear_freq_refined_k4)

print(comparison, simplify = FALSE)
rm(fit_no_slopes)
rm(fit_k3)
rm(fit_full)
rm(linear_freq)
rm(refined_fit_k4)
rm(linear_freq_refined_k4)
```


# Based on model comparison, seems like there is no major difference between k=3 and k=10 (k=10 is the "full" model), so I'll pick the simpler model. On top of that, we can have a non-linear interaction of frequency with time or a linear one (the one with prefix `linear_freq`). The latter model shows an intuitive fixed effect of frequency, even though I still need to check how to best interpret time and frequency interaction.

```{r}
full.fit <- readRDS("models/refined_model_brms_fit_k4_linear_freq.rds")
summary(full.fit)
```

The fixed effect of frequency is actually non-overlapping with zero for its 95% CI:
```{r}
library(bayesplot)
library(ggplot2)

# Plot posterior distributions for fixed effects (Intercept, bipartite, element_type)
# We exclude the spline coefficients (starting with "s") to keep the plot readable
mcmc_plot(full.fit, 
          type = "areas", 
          prob = 0.95, 
          pars = "^b_") +
  ggtitle("Posterior Distributions of Fixed Effects")
```

```{r}
conditional_effects(full.fit, 
                    effects = "date:is_bipartite", 
                    points = TRUE) # points=TRUE overlays the raw data (optional but helpful)


# dy_bipartite/dt - dy_non-bipartite/dt plotting (next cell)
```

There seems to be no support for Paul's proposal and at later dates, the reverse seems to be true according to this model.

```{r}
library(tidybayes)
library(modelr)


# 1. Generate a comprehensive grid for all permutations
grid_full <- model_data %>%
  data_grid(
    date = seq_range(date, n = 80), 
    is_bipartite = c("0", "1"),
    element_type = c("vowel", "consonant"),
    variety = c("Central German", "Upper German"),
    log_freq = 0 # Held at mean
  ) 

# 2. Extract posterior draws and compute the difference (Bipartite - Non-Bipartite)
# We use 'epred' for the difference in expected probability
plot_data <- grid_full %>%
  add_epred_draws(full.fit, re_formula = NA) %>%
  ungroup() %>%
  select(date, is_bipartite, element_type, variety, .epred, .draw) %>%
  pivot_wider(names_from = is_bipartite, values_from = .epred) %>%
  mutate(diff = `1` - `0`)

# 3. Create the 4-panel faceted plot
ggplot(plot_data, aes(x = date, y = diff)) +
  # Add the 95% and 80% credible intervals
  stat_lineribbon(.width = c(.95, .80), alpha = 0.6) +
  # Reference line at zero
  geom_hline(yintercept = 0, linetype = "dashed", color = "red") +
  # Facet by Variety and Element Type
  facet_grid(variety ~ element_type) +
  scale_fill_brewer(name = "Credible Interval") +
  labs(
    title = "Effect of Bipartite Marking on Leveling Probability",
    subtitle = "Difference: Bipartite minus Non-Bipartite (Negative values support Paul's Principle)",
    x = "Year",
    y = "Difference in Probability"
  ) +
  theme_minimal() +
  theme(panel.spacing = unit(2, "lines"), 
        legend.position = "bottom")
```

The effect of time is clear and uncontroversial, there is steady increase in levelling probability over time. 

```{r}
conditional_effects(full.fit, 
                    effects = "date")
```

What we discussed, consonants seem to be more likely to be levelled than vowels. If we look at the fixed effect of vowel vs consonant, this fixed effect also doesn't overlap with 0 for 95% CI. 

```{r}

conditional_effects(full.fit, 
                    effects = "date:element_type", 
                    )
```


Overall, plural seems to be more likely to be levelled:
```{r}
mcmc_plot(full.fit, 
          type = "intervals", 
          regex = TRUE, 
          variable = "^r_std_infl") +
  geom_vline(xintercept = 0, linetype = "dashed", color = "red") +
  labs(title = "Effect of Inflectional Context (Deviations)",
       subtitle = "Positive = More prone to leveling | Negative = More resistant",
       x = "Deviation from Global Intercept (Log-Odds)")
```
